The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-15 paper/ms.tex
finished section 2

\newcommand{\Gus}[1]{\textcolor{red}{#1}}

@@

The upper right panel of Figure~\ref{fig:many_orbit_wrong_ref} shows that the
fractional error in $J_z$ is more exaggerated for more planar (disk-like)
orbits. For \thin{} (\thick{}), a systematic offset of $15\,\pc$ ($120\,\pc$) in the $z$-coordinate results in a $25\%$ deviation in $J_z$.
We find that \halo{} is relatively resistant to errors in the midplane, with
only $\sim15\%$ error in $J_z$ out to an offset of $500\,\pc$.

@@

In each panel of Figure~\ref{fig:many_orbit_wrong_ref}, where relevant, we
include the estimation of the action errors derived under the epicyclic
approximation from Equation~\eqref{eq:Ji_err_mosttime}, with $\Delta
v_{\phi}=0$, as dashed lines in the color of each orbit. This equation is
relevant since during most of the orbit the star will be at maximum radial and
vertical amplitude. Note that we
consider an error in the $x$-coordinate $\Delta x$, which is not exactly the
same as $\Delta R$. For observations of stars close to us, we have that
$\Delta x \sim \Delta R$, but for the experiment performed in this section we
consider observations of the star throughout its entire orbit. This introduces
a factor of $2/\pi$ when converting from $\Delta x$ to $\Delta R$, which we
derive in Appendix~\ref{app:deltax}.

@@

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, teal, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For each orbit, we set the initial position to be
$(8,0,0)\,\kpc$ and the initial velocity to be $(0, -190, v_z)\,\kms$, where
we vary $v_z$.\footnote{\label{note:vz_orbits}One can recover \thin{}, \thick{}, and \halo{} by setting $v_z=10$, $50$, and $190$, respectively (see Table~\ref{tab:orbits}).} For a \thin{}-like orbit ($J_z\sim0.5\,\actunit$), even a
$10\,\pc$ offset in $z$ is enough to induce a $\sim25\%$ error in $J_z$. For
larger values of $J_z$, the fractional errors are suppressed, but the induced
error can still be large depending on how great the $z$ offset is. We also
plot the epicylic prediction for $\Delta J_z / J_z$ from
Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$ offset. We
find that the epicyclic approximation matches the numerical estimate quite
well.

>>>>

\newcommand{\Gus}[1]{\textcolor{red}{#1}}
\newcommand{\MM}[1]{\textcolor{blue}{#1}}   %mm  Mordecai

@@

The upper right panel of Figure~\ref{fig:many_orbit_wrong_ref} shows that the
fractional error in $J_z$ is more exaggerated for more planar (disk-like)
orbits. For \thin{}, 
%mm (\thick{}), 
a systematic offset of $15\,\pc$
%mm ($120\,\pc$)
in the $z$-coordinate results in a $25\%$ deviation in $J_z$,
%mm
     while a $120\,\pc$ offset results in the same deviation for \thick{}.
We find that \halo{} is relatively resistant to errors in the midplane, with
only $\sim15\%$ error in $J_z$ out to an offset of $500\,\pc$.

@@

In each panel of Figure~\ref{fig:many_orbit_wrong_ref}, where relevant, we
include the estimation of the action errors derived under the epicyclic
approximation from Equation~\eqref{eq:Ji_err_mosttime}, with $\Delta
v_{\phi}=0$, as dashed lines in the color of each orbit. This equation is
relevant since during most of the orbit the star will be 
%mm at 
    close to 
maximum radial and
vertical amplitude. Note that we
consider an error in the $x$-coordinate $\Delta x$, which is not exactly the
same as $\Delta R$. For observations of stars close to us, we have that
$\Delta x \sim \Delta R$, but for the experiment performed in this section we
consider observations of the star throughout its entire orbit. This introduces
a factor of $2/\pi$ when converting from $\Delta x$ to $\Delta R$, which we
derive in Appendix~\ref{app:deltax}.

@@

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of 
%mm $10, 50,
     10, 50, and $
100\,\pc$ (orange, teal, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. 
For each orbit, we set the initial position to be
$(8,0,0)\,\kpc$ and the initial velocity to be $(0, -190, v_z)\,\kms$, where
we vary $v_z$.\footnote{\label{note:vz_orbits}One can recover \thin{},
  \thick{}, and \halo{} by setting $v_z=10$, $50$, and $190
%mm
     \kms$,
 respectively, 
%mm [one significant figure values]
      giving $J_z \simeq 0.7, 20,$ and $500\,\actunit$  
(see Table~\ref{tab:orbits}).} For a \thin{}-like orbit ($J_z\sim
%mm 0.5
         0.7%
\,\actunit$), 
even a
$10\,\pc$ offset in $z$ is enough to induce a 
%mm $\sim25\%$ 
         $\sim20\%$ 
error in $J_z$. For
larger values of $J_z$, the fractional errors are suppressed, but the induced
error can still be large depending on how great the $z$ offset is. We also
plot the epicylic prediction for $\Delta J_z / J_z$ from
Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$ offset. We
find that the epicyclic approximation matches the numerical estimate quite
well.
